Extended Data

### 2022-03-08 doc/manuscript/Dow_manuscript_Nature.Rmd
Update Dow_manuscript_Nature.Rmd

In recent decades, tree growth in Earth's forests has more than offset losses from deforestation and other disturbances, such that a net forest CO~2~ sink of ~`r round(5.8*12/44,1)` Gt C yr^-1^ <!--net forest estimate in Harris et al. Table 2--> offsets ~`r round(5.8/32,1)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al. Table 2)--> of anthropogenic emissions[@harris_global_2021], dramatically slowing the pace of atmospheric CO~2~ accumulation and climate change.
Of this important C sink, ~`r round(3.6/7.6,2)*100`%<!-- temperate net / global net, (Harris et al. Table 1)--> occurs in temperate forests[@harris_global_2021], with temperate deciduous forests sequestering >0.3 Gt C yr^-1^ [@pugh_role_2019]. <!-- Pugh et al. 2019 estimate that the temperate deciduous sink is >300 Tg C yr-1, with ~200 Tg C yr-1  attributable to demography (forest regrowth) and >100 Tg C yr-1  controlled by environmental drivers.--> 
The future behavior of this C sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020;@friedlingstein_global_2020].

@@

We have little evidence as to how stem growth and $ANPP_{woody}$ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016]. 
Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] and increased annual growth in mesic climates[@gao_earlier_2022;@anderson-teixeira_joint_2021; @more_conifer_refs]. 
Much less is known about how deciduous broadleaf tree growth responds to spring temperatures.
While close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the timing of growth earlier alongside leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020], it has been observed that annual growth is more dependent on conditions within the growing season then on growing season length[@etzold_number_2022]. 
Furthermore, recent evidence suggests that growth of broadleaf deciduous trees may be sink-saturated, such that longer growing seasons with more C fixation do not necessarily augment growth[@refs].
Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest, but usually insignificant, positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms[@refs].

@@

<!--These following two paragraphs have a lot of overlap and need work. They should probably be merged into one. -->
In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
Current terrestrial ecosystem models, which have a rudimentary treatment of tree growth phenology[@delpierre_temperate_2016], effectively assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
However, recent experimental and observational findings show that annual stem growth is not correlated with growing season length[@etzold_number_2022], *but rather can be limited by sink factors[@keenan_timing_2015;@zani_increased_2020;@zohner_how_2021;@delpierre_wood_2016]*,<!--I (KAT) need to work on the part in italics, if retained keenan_timing and zani_increased_2020 do not look at stem growth and are largely contradicted by lu_no_2022.--> and that positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018]. 
These studies suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.

>>>>

In recent decades, tree growth in Earth's forests has more than offset losses from deforestation and other disturbances, such that a net forest CO~2~ sink of ~`r round(5.8*12/44,1)` Gt C yr^-1^<!--net forest estimate in Harris et al. Table 2--> offsets ~`r round(5.8/32,1)*100`%<!--net forest estimate / fossil fuel & cement, (Harris et al. Table 2)--> of anthropogenic emissions[@harris_global_2021], dramatically slowing the pace of atmospheric CO~2~ accumulation and climate change.
Of this important C sink, ~`r round(3.6/7.6,2)*100`%<!-- temperate net / global net, (Harris et al. Table 1)--> occurs in temperate forests[@harris_global_2021], with temperate deciduous forests sequestering >0.3 Gt C yr^-1^ [@pugh_role_2019]. <!-- Pugh et al. 2019 estimate that the temperate deciduous sink is >300 Tg C yr-1, with ~200 Tg C yr-1  attributable to demography (forest regrowth) and >100 Tg C yr-1  controlled by environmental drivers.--> 
The future behavior of this C sink will play an important yet uncertain role in influencing atmospheric CO~2~ and climate change[@arora_carbon_2020;@friedlingstein_global_2020].

@@

We have little evidence as to how stem growth and $ANPP_{woody}$ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016]. 
Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;@delpierre_chilling_2019;@rossi_critical_2008] and increased annual growth in mesic climates[@gao_earlier_2022;@anderson-teixeira_joint_2021; @more_conifer_refs]. 
Much less is known about deciduous tree responses[@savage_coordination_2021;@delpierre_temperate_2016].
Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020;@stuble_plant_2021;@delgado_differences_2020].
However, earlier growth would not necessarily translate to earlier, faster, or more stem growth; rather, cell enlargement is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022]. 
Furthermore, *growth of broadleaf deciduous trees may be sink-saturated*, such that longer growing seasons with more C fixation do not necessarily augment growth[@refs].
Tree-ring records, which can be used to examine relationships of annual growth to temperature but not to understand growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest, but usually insignificant, positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014;@lu_no_2022].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms[@refs].

@@

<!--These following two paragraphs have a lot of overlap and need work. They should probably be merged into one. -->
~~In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
Current terrestrial ecosystem models, which have a rudimentary treatment of tree growth phenology[@delpierre_temperate_2016], effectively assume that longer growing seasons lead to increasing annual net CO~2~ uptake (i.e., net ecosystem exchange, _NEE_)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].~~
However, recent experimental and observational findings show that annual stem growth is not correlated with growing season length[@etzold_number_2022], *but rather can be limited by sink factors[@keenan_timing_2015;@zani_increased_2020;@zohner_how_2021;@delpierre_wood_2016]*,<!--I (KAT) need to work on the part in italics, if retained keenan_timing and zani_increased_2020 do not look at stem growth and are largely contradicted by lu_no_2022.--> and that positive effects of warm springs can be compensated by negative effects of accumulation of seasonal water deficits[@buermann_widespread_2018]. 
~~These studies suggest that warmer spring temperatures may not have the expected positive effect on forest CO~2~ sequestration.~~
